---
title: "Nov-1-2024-practice.qmd"
format: html
editor: visual
---

## Introduction

At this point, you have learned to perform and interpret three different statistical tests: Chi-square test of association (contingency analysis), two-sample t-test and linear regression.

Now it is time for you to practice what you've been learning. Please add the data set called "A_morio_2014_plants.csv" to your Biostats-In-Class RStudio project as you will need it for this exercise. The data are associated with [this paper in the journal Evolution](https://academic.oup.com/evolut/article/70/3/716/6852093?login=true) and come from [this repository on Dryad](https://datadryad.org/stash/dataset/doi:10.5061/dryad.nf663). The authors of the study were interested in the degree to which pollinators apply selective pressure on a variety of flower traits in a species of orchid plant. In the experiment, some of the orchids were allowed to experience natural pollination and were considered the control group, and another group of orchids was pollinated by hand by the researchers (the "hand-pollinated" treatment group).
```{r}
rm(list = ls())
library(here)
library(tidyverse)
library(ggfortify)
morio<-read.csv(here("Data", "A_morio_2014_plants.csv"))
```


### Question 1

The researchers wished to know whether spur length (the "spur" is a part of the flower that sticks out) of flowers differed between control and hand-pollinated plants.

Your task is to determine the appropriate statistical test you should use to answer this question from the data and then proceed with the statistical analysis, using the "plot your data (and guess the answer!) -\> build your model -\> check your assumptions -\> interpret your model" workflow (for now, we will leave out the last step which is to make a better plot at the end).

I am expecting that you will document your process clearly along the way.

The researchers wished to know if there was an association between plant height class and treatment group and the number of flowers produced by orchid plants.

Your task is to determine the appropriate statistical test you should use to answer this question from the data and then proceed with the statistical analysis, using the "plot your data (and guess the answer!) -\> build your model -\> check your assumptions -\> interpret your model" workflow (for now, we will leave out the last step which is to make a better plot at the end).

I am expecting that you will document your process clearly along the way.
```{r}
#clean up data
glimpse(morio)
morio$treatment<- as.factor(morio$treatment)
morio$plant_ht_class<- as.factor(morio$plant_ht_class)
```

```{r}
#make dataset
q1<- morio |> select(treatment, spur_length_mm)
q1
```

type of data
spure length = response and continuous 
treatment = predictor categorical 

test to run is a t test
```{r}
ggplot(morio, aes(x=spur_length_mm))+
  geom_histogram()+
  facet_wrap(~treatment)
```
Assumptions
- data looks pretty normal and equal variance has been met 
- looking to see if there is a difference between group and if there is it is small
- peaks are both at about 10

```{r}
t.test(spur_length_mm ~ treatment, data = morio)
```

- 95% CI does not include 0 indicates there is a difference between groups
Reject the null hypothesis that there is no significant difference in spur length between hand pollinated and control as the p-value is less than 0.5

### Question 2

The researchers wished to know if there was an association between plant height class and treatment group and the number of flowers produced by orchid plants.

Your task is to determine the appropriate statistical test you should use to answer this question from the data and then proceed with the statistical analysis, using the "plot your data (and guess the answer!) -\> build your model -\> check your assumptions -\> interpret your model" workflow (for now, we will leave out the last step which is to make a better plot at the end).

I am expecting that you will document your process clearly along the way.
```{r}
#makedata set
q2<- morio |> select(plant_ht_class, treatment, num_flowers)
q2
```

response is number of flowers produced by orchid plants
predictors are plant height class and treatment group
chi square goodness of fit test

```{r}
#organize data for plotting and analysis
totals<- morio |> group_by(plant_ht_class, treatment)|>
  summarise(total.numer = sum(num_flowers))
totals
```

```{r}
#plot
ggplot(totals, aes(x=plant_ht_class, y= total.numer, fill = treatment))+
  geom_bar(stat = 'identity', position = 'dodge')
```

- looks like there is a different between short and tall plants whether they were the control or hand pollinated 

```{r}
lady.mat<- xtabs(num_flowers ~ plant_ht_class + treatment, data = morio)
```

```{r}
chisq.test(lady.mat)
```

reject alternative hypothesis as p-value is greater than 0.05 there is no association between plant height class and treatment group and the number of flowers produced by orchid plants.

### Question 3

The researchers wished to know if there is an association between corolla area and spur length in hand-pollinated flowers. 

both are continuous predictors are seeing there is a association so will be preforming a regression 

```{r}
q3<- morio |> filter(treatment == "hand-pollinated")
q3
```

```{r}
ggplot(q3, aes(x=corolla_area_mm2, y=spur_length_mm))+
  geom_point()
```

```{r}
model_pgr<- lm(spur_length_mm ~corolla_area_mm2, data = q3)
```


```{r}
autoplot(model_pgr, smooth.colour = NA)
#all assumptions look good
```

```{r}
anova(model_pgr)
```

```{r}
summary(model_pgr)
```

reject null hypothesis that there is no association between plant height class and treatment group and the number of flowers produced by orchid plants


Your task is to determine the appropriate statistical test you should use to answer this question from the data and then proceed with the statistical analysis, using the "plot your data (and guess the answer!) -\> build your model -\> check your assumptions -\> interpret your model" workflow (for now, we will leave out the last step which is to make a better plot at the end).

I am expecting that you will document your process clearly along the way.
