---
title: "30 October 2024 - Regression"
format: html
editor: visual
---

## Overview

For your homework you read through the section of the book that presents linear regression. Regression is a powerful tool that is used widely in Biology so let's get more familiar with it by running another example.

Set things up (hint: there is something new to add to your `library()` calls...)

```{r}
rm(list=ls())
library(here)
library(tidyverse)
library(ggfortify)
```

## Linear models

Now we move into something called "general linear models". They are a class of model that includes regression, multiple regression, ANOVA and ANCOVA. All are fundamentally linear models.

Don't confuse a general linear model with a *generalized* linear model or GLM.

## The data

Let's pull in some data appropriate for the day before Halloween: data on Bats!

```{r}
bats<- read.csv(here("Data","tricolored_bat_hibernation.csv"))
glimpse(bats)
```

These data came from a research project looking at what, if any, environmental variables impact mass loss in bats during hibernation. Bats were housed in environmental chambers (`Chamber`) with different temperature (`Temp_C`) and humidity (`Rel_humid_percent`) levels and were monitored for the number of different times they aroused from hibernation as well as how much body mass they lost during the hibernation period.

A couple of the variables are categorical - let's fix that in the data.

```{r}
bats$Sex <- as.factor(bats$Sex)
bats$Chamber<- as.factor(bats$Chamber)
```

## Question

Is there a relationship between the number of arousals and the amount of mass lost in bats?

Prediction: bats that arouse from hibernation more frequently will lose more body mass during hibernation.

## Plot the data

Let's start with a scatter plot:

```{r}
ggplot(bats, aes(x=Arousals, y=mass_loss_g))+
  geom_point()
```

## Interpreting the figure: biological insight

We notice that the slope is: **CHOOSE ONE: positive **

We can also guesstimate the slope and intercept from the figure:

Roughly speaking, the mass lost during hibernation ranges between **0.2 and 1.9**

Roughly speaking, the number of arousals varies between **4 to 16**

Slope is rise/run, which in this case is about **1.7** for rise and **12** for run, so we get **0.146** as our guesstimate of the slope.

Eyeballing the plot, we would also guess that the y-intercept is somewhere around **0.1**

**IT IS ALWAYS A GOOD IDEA TO EXAMINE YOUR DATA BEFORE THE ANALYSIS AND TO SEE IF YOU CAN APPROXIMATE SOME OF THE VALUES THAT WILL RESULT.**

## Making a simple linear regression happen

We use the `lm()` function...

```{r}
model_bats <- lm(Arousals ~ mass_loss_g, data = bats)
```

Ok, the model has been fit. Before we look at the results, let's consider the assumptions.
- top let: equal variance is good, a linear model is good (linear line is good to fit this)
- top right: normality is good
- bottom left: equal variance is good 
- bottom right: leverage is good 

```{r}
autoplot(model_bats, smooth.colour = NA)
```

What does this all mean?

## Now the interpretation

```{r}
anova(model_bats)
```

And the summary table:

```{r}
summary(model_bats)
```

## From stats back to figure

Now let's make a figure that shows off our relationship in light of our statistical results.

```{r}
ggplot(bats, aes(x= Arousals, y= mass_loss_g))+
  geom_point()+
  geom_smooth(method = 'lm')
```

We can see that the `geom_smooth()` call allowed the computer to add the fitted values and the standard error of the fit to a figure.

Don't expect `geom_smooth()` to work correctly for more complex models.
