---
title: "rabies.qmd"
format: html
editor: visual
---

## Rabies project

You are receiving a dataset from the New York State Department of Health that shows test results for specimens submitted to be tested for rabies.

## In Class Exercise 1

### Set things up and read in the data

Add the code to set up the workspace

```{r}
rm(list=ls())
library(here)
library(tidyverse)
```

Now read in the data and name it as an object called `rabies`.

```{r}
rabies<- read.csv(here("Data", "Rabies_Laboratory_Submissions.csv"))

```

Reduce the dataset to just the first 7 columns

```{r}
rabies<- rabies |> select(1:7)
```

### Examine the data

1.  Get a summary for each variable.

```{r}
summary(rabies)
names(rabies)
```

2.  Repair column names

```{r}
names(rabies)  <- c("Received_Year", "Received_Month", "Report_Date",         "County_Name","Animal_Classification", "Number_of_Samples",     "Positive_Samples" )
```

3.  Adjust data types

```{r}
rabies$County_Name<- as.factor(rabies$County_Name)
rabies$Animal_Classification <- as.factor(rabies$Animal_Classification)
rabies$Report_Date <- mdy(rabies$Report_Date)
levels(rabies$County_Name)
levels(rabies$Animal_Classification)
```

4.  Now plot some data

```{r}
ggplot(rabies, aes(Received_Year))+
  geom_histogram(bins = 15)

ggplot(rabies, aes(Animal_Classification))+
  geom_bar()
ggplot(rabies, aes(Animal_Classification))+
  geom_histogram(stat = "count")
```

## In Class Exercise 2

### Look at some relationships

5.  Create a percent positive column

```{r}
rabies <- rabies |> 
  mutate(Percent_pos = Positive_Samples / Number_of_Samples * 100)
```

6.  Using the rabies dataset, create a new object called "animal summary" in which you determine, by Year and Animal Classification, the total number of samples, total positive samples, and the mean percent positive samples.

```{r}
animal_summary <- rabies |>
  group_by(Received_Year, Animal_Classification) |>
  summarise(
  Number_Samples = sum(Number_Samples),
  Positive_Samples = sum(Positive_Samples),
  mean_percent_pos = mean(percent_pos)
)
```

Next, make a boxplot of those data with the following features: - The x-axis should be the animal classificaton - The y-axis should be the mean percent positive samples - Your boxplot should include the individual points laid over the boxplots. Make the points gray. - Rename the x-axis label as "Animal type" and the y-axis label as "Mean percent positive samples"

```{r}
ggplot(animal_summary, aes(x = Animal.Classification, y = mean_percent_pos)) +
  geom_boxplot() +
  geom_jitter(color = "gray") + 
  labs(
    x = "Animal type",
    y = "Mean percent positive samples"
  ) +
  theme_minimal()
```
